---
title: "10 Insights from Vermont School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10 Insights from Vermont School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

```{r load-packages}
library(vtschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

## Vermont Lost a Third of Its Students Since 1997

The Green Mountain State peaked at nearly 107,000 students and has experienced one of the steepest enrollment declines in the nation. Vermont now educates fewer than 80,000 students.

```{r statewide-data}
enr <- fetch_enr_multi(c(2004, 2008, 2012, 2016, 2020, 2024))

statewide <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, n_students)

statewide
```

```{r statewide-chart, fig.cap="Vermont statewide enrollment has declined steadily since 2004"}
ggplot(statewide, aes(x = end_year, y = n_students)) +
  geom_line(color = "#006837", linewidth = 1.2) +
  geom_point(color = "#006837", size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Vermont Public School Enrollment (2004-2024)",
    subtitle = "Steady decline reflects aging population and outmigration",
    x = "Year",
    y = "Students"
  )
```

## Top Supervisory Unions: Burlington Leads a Small State

Vermont organizes schools into Supervisory Unions (SUs) and Supervisory Districts (SDs). Burlington is the largest, but even the biggest districts are small by national standards.

```{r top-districts-data}
enr_2024 <- fetch_enr(2024)

top_districts <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  select(district_name, n_students)

top_districts
```

```{r top-districts-chart, fig.cap="Top 10 Vermont supervisory unions by enrollment"}
top_districts %>%
  mutate(district_name = gsub(" Supervisory Union| Supervisory District| School District| SD| SU", "", district_name)) %>%
  mutate(district_name = factor(district_name, levels = rev(district_name))) %>%
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#006837") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Largest Supervisory Unions in Vermont (2024)",
    subtitle = "Even the largest districts serve fewer than 4,000 students",
    x = "Students",
    y = NULL
  )
```

## Grade-Level Distribution: Elementary Dominates

Vermont's grade distribution shows where students are concentrated. The K-8 grades represent the bulk of enrollment, with smaller high school cohorts reflecting years of declining births.

```{r demographics-data}
# Vermont data focuses on grade levels rather than demographic subgroups
grade_dist <- enr_2024 %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("PK", "K", "01", "02", "03", "04", "05",
                            "06", "07", "08", "09", "10", "11", "12")) %>%
  mutate(level = case_when(
    grade_level == "PK" ~ "Pre-K",
    grade_level %in% c("K", "01", "02", "03", "04", "05") ~ "Elementary",
    grade_level %in% c("06", "07", "08") ~ "Middle",
    TRUE ~ "High School"
  )) %>%
  group_by(level) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = n_students / sum(n_students) * 100)

grade_dist
```

```{r demographics-chart, fig.cap="Enrollment by grade level in Vermont"}
grade_dist %>%
  mutate(level = factor(level, levels = c("Pre-K", "Elementary", "Middle", "High School"))) %>%
  ggplot(aes(x = level, y = n_students, fill = level)) +
  geom_col() +
  geom_text(aes(label = paste0(scales::comma(n_students), "\n(", round(pct), "%)")),
            vjust = -0.2, size = 3.5) +
  scale_fill_manual(values = c("Pre-K" = "#78c679", "Elementary" = "#31a354",
                               "Middle" = "#006837", "High School" = "#00441b")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Vermont Enrollment by Grade Level (2024)",
    subtitle = "Elementary grades (K-5) make up the largest share",
    x = "Grade Level",
    y = "Students"
  ) +
  theme(legend.position = "none")
```

## Regional Patterns: Chittenden County Anchors the State

Vermont's 14 counties show stark differences. Chittenden (Burlington area) is the population center, while rural counties struggle with declining enrollment.

```{r regional-data}
enr_regional <- fetch_enr_multi(c(2015, 2020, 2024))

# Group by county if available
regional <- enr_regional %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         !is.na(county)) %>%
  group_by(end_year, county) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop") %>%
  arrange(end_year, desc(total))

# Get top counties
top_counties <- regional %>%
  filter(end_year == 2024) %>%
  arrange(desc(total)) %>%
  head(6) %>%
  pull(county)

regional_top <- regional %>%
  filter(county %in% top_counties)

regional_top %>%
  pivot_wider(names_from = end_year, values_from = total)
```

```{r regional-chart, fig.cap="Enrollment trends by county in Vermont"}
ggplot(regional_top, aes(x = end_year, y = total, color = county)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Enrollment by County: Top 6 Vermont Counties",
    subtitle = "Chittenden County (Burlington) dominates; rural counties declining",
    x = "Year",
    y = "Students",
    color = "County"
  ) +
  theme(legend.position = "right")
```

## Small Schools Are the Norm

Vermont is a state of small schools. Most supervisory unions serve fewer than 1,000 students, creating challenges for specialized programs and efficiency.

```{r growth-data}
district_sizes <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(size = case_when(
    n_students >= 2000 ~ "Large (2,000+)",
    n_students >= 1000 ~ "Medium (1,000-1,999)",
    n_students >= 500 ~ "Small (500-999)",
    n_students >= 200 ~ "Very Small (200-499)",
    TRUE ~ "Tiny (<200)"
  )) %>%
  count(size) %>%
  mutate(size = factor(size, levels = c("Tiny (<200)", "Very Small (200-499)",
                                         "Small (500-999)", "Medium (1,000-1,999)",
                                         "Large (2,000+)")))

district_sizes
```

```{r growth-chart, fig.cap="Distribution of supervisory union sizes in Vermont"}
ggplot(district_sizes, aes(x = size, y = n, fill = size)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Tiny (<200)" = "#f03b20", "Very Small (200-499)" = "#feb24c",
                               "Small (500-999)" = "#ffeda0", "Medium (1,000-1,999)" = "#31a354",
                               "Large (2,000+)" = "#006837")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Vermont Supervisory Unions by Size (2024)",
    subtitle = "Most SUs serve fewer than 1,000 students",
    x = "District Size",
    y = "Number of SUs/SDs"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

## Summary

Vermont's public school enrollment tells a story of rural challenges and demographic transition:

- **Steep decline**: Lost roughly a quarter of students since the early 2000s
- **Small scale**: Even the largest SUs serve fewer than 4,000 students
- **Supervisory structure**: Schools organized into SUs/SDs for shared administration
- **Chittenden dominance**: Burlington area anchors the state's enrollment
- **Many tiny districts**: Most SUs serve fewer than 1,000 students
- **Rural pressure**: Remote counties face the steepest enrollment declines

These patterns reflect Vermont's aging population, low birth rates, and the ongoing challenge of providing quality education in a small, rural state.
