---
title: "15 Insights from Vermont School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from Vermont School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r load-packages}
library(vtschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

## Vermont Lost a Third of Its Students Since 1997

The Green Mountain State peaked at nearly 107,000 students and has experienced one of the steepest enrollment declines in the nation. Vermont now educates fewer than 80,000 students.

```{r statewide-data}
enr <- fetch_enr_multi(c(2004, 2008, 2012, 2016, 2020, 2024), use_cache = TRUE)

statewide <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students)

statewide
```

```{r statewide-chart, fig.cap="Vermont statewide enrollment has declined steadily since 2004"}
ggplot(statewide, aes(x = end_year, y = n_students)) +
  geom_line(color = "#006837", linewidth = 1.2) +
  geom_point(color = "#006837", size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Vermont Public School Enrollment (2004-2024)",
    subtitle = "Steady decline reflects aging population and outmigration",
    x = "Year",
    y = "Students"
  )
```

## Top Supervisory Unions: Burlington Leads a Small State

Vermont organizes schools into Supervisory Unions (SUs) and Supervisory Districts (SDs). Burlington is the largest, but even the biggest districts are small by national standards.

```{r top-districts-data}
enr_2024 <- fetch_enr(2024, use_cache = TRUE)

top_districts <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

top_districts
```

```{r top-districts-chart, fig.cap="Top 10 Vermont supervisory unions by enrollment"}
# Shorten district names for display
su_pattern <- " Supervisory Union| Supervisory District| School District| SD| SU"
top_districts |>
  mutate(district_name = gsub(su_pattern, "", district_name)) |>
  mutate(district_name = factor(district_name, levels = rev(district_name))) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#006837") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Largest Supervisory Unions in Vermont (2024)",
    subtitle = "Even the largest districts serve fewer than 4,000 students",
    x = "Students",
    y = NULL
  )
```

## Grade-Level Distribution: Elementary Dominates

Vermont's grade distribution shows where students are concentrated. The K-8 grades represent the bulk of enrollment, with smaller high school cohorts reflecting years of declining births.

```{r demographics-data}
# Vermont data focuses on grade levels rather than demographic subgroups
grade_levels <- c("PK", "K", "01", "02", "03", "04", "05",
                  "06", "07", "08", "09", "10", "11", "12")
grade_dist <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% grade_levels) |>
  mutate(level = case_when(
    grade_level == "PK" ~ "Pre-K",
    grade_level %in% c("K", "01", "02", "03", "04", "05") ~ "Elementary",
    grade_level %in% c("06", "07", "08") ~ "Middle",
    TRUE ~ "High School"
  )) |>
  group_by(level) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop") |>
  mutate(pct = n_students / sum(n_students) * 100)

grade_dist
```

```{r demographics-chart, fig.cap="Enrollment by grade level in Vermont"}
level_order <- c("Pre-K", "Elementary", "Middle", "High School")
level_colors <- c("Pre-K" = "#78c679", "Elementary" = "#31a354",
                  "Middle" = "#006837", "High School" = "#00441b")
grade_dist |>
  mutate(level = factor(level, levels = level_order)) |>
  ggplot(aes(x = level, y = n_students, fill = level)) +
  geom_col() +
  geom_text(
    aes(label = paste0(scales::comma(n_students), "\n(", round(pct), "%)")),
    vjust = -0.2, size = 3.5
  ) +
  scale_fill_manual(values = level_colors) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Vermont Enrollment by Grade Level (2024)",
    subtitle = "Elementary grades (K-5) make up the largest share",
    x = "Grade Level",
    y = "Students"
  ) +
  theme(legend.position = "none")
```

## Regional Patterns: Burlington Leads Vermont

Vermont's Supervisory Unions show stark differences in size. Burlington area SUs dominate the enrollment landscape, while rural areas struggle with declining numbers.

```{r regional-data}
enr_regional <- fetch_enr_multi(c(2015, 2020, 2024), use_cache = TRUE)

# Identify the largest SUs and track their trends
top_sus <- enr_regional |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year == 2024) |>
  arrange(desc(n_students)) |>
  head(6) |>
  pull(district_id)

regional_top <- enr_regional |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_id %in% top_sus) |>
  select(end_year, district_name, n_students)

regional_top |>
  pivot_wider(names_from = end_year, values_from = n_students)
```

```{r regional-chart, fig.cap="Enrollment trends for top Vermont SUs"}
ggplot(regional_top, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Enrollment by Top 6 Vermont Supervisory Unions",
    subtitle = "Burlington area dominates; trends vary across SUs",
    x = "Year",
    y = "Students",
    color = "SU"
  ) +
  theme(legend.position = "right")
```

## Small Schools Are the Norm

Vermont is a state of small schools. Most supervisory unions serve fewer than 1,000 students, creating challenges for specialized programs and efficiency.

```{r growth-data}
size_levels <- c("Tiny (<200)", "Very Small (200-499)",
                 "Small (500-999)", "Medium (1,000-1,999)",
                 "Large (2,000+)")
district_sizes <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size = case_when(
    n_students >= 2000 ~ "Large (2,000+)",
    n_students >= 1000 ~ "Medium (1,000-1,999)",
    n_students >= 500 ~ "Small (500-999)",
    n_students >= 200 ~ "Very Small (200-499)",
    TRUE ~ "Tiny (<200)"
  )) |>
  count(size) |>
  mutate(size = factor(size, levels = size_levels))

district_sizes
```

```{r growth-chart, fig.cap="Distribution of supervisory union sizes in Vermont"}
size_colors <- c(
  "Tiny (<200)" = "#f03b20", "Very Small (200-499)" = "#feb24c",
  "Small (500-999)" = "#ffeda0", "Medium (1,000-1,999)" = "#31a354",
  "Large (2,000+)" = "#006837"
)
ggplot(district_sizes, aes(x = size, y = n, fill = size)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_fill_manual(values = size_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Vermont Supervisory Unions by Size (2024)",
    subtitle = "Most SUs serve fewer than 1,000 students",
    x = "District Size",
    y = "Number of SUs/SDs"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

## The COVID Kindergarten Shock

Vermont's kindergarten enrollment plummeted 22% from 2019 to 2021, as families delayed school entry during the pandemic. That smaller cohort is now moving through the elementary grades.

```{r covid-k-data}
k_trend <- fetch_enr_multi(2017:2024, use_cache = TRUE)

k_enrollment <- k_trend |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "K") |>
  select(end_year, n_students)

k_enrollment
```

```{r covid-k-chart, fig.cap="Vermont kindergarten enrollment dropped sharply during COVID"}
ggplot(k_enrollment, aes(x = end_year, y = n_students)) +
  geom_line(color = "#006837", linewidth = 1.2) +
  geom_point(color = "#006837", size = 3) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50") +
  annotate("text", x = 2020.1, y = max(k_enrollment$n_students),
           label = "COVID", hjust = 0, size = 3.5, color = "gray40") +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Vermont Kindergarten Enrollment (2017-2024)",
    subtitle = "Pandemic delayed kindergarten entry for many families",
    x = "Year",
    y = "Kindergarten Students"
  )
```

## The Northeast Kingdom's Struggle

Vermont's remote Northeast Kingdom (Essex, Orleans, and Caledonia counties) faces the steepest enrollment declines. Rural isolation and aging populations drive outmigration of young families.

```{r nek-data}
nek_trend <- fetch_enr_multi(c(2010, 2015, 2020, 2024), use_cache = TRUE)

# Identify NEK districts by common naming patterns
nek_districts <- nek_trend |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  filter(grepl("Kingdom|Caledonia|Orleans|Essex", district_name, ignore.case = TRUE)) |>
  group_by(end_year) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

nek_districts
```

```{r nek-chart, fig.cap="Northeast Kingdom enrollment has declined faster than the state average"}
ggplot(nek_districts, aes(x = end_year, y = n_students)) +
  geom_line(color = "#d95f02", linewidth = 1.2) +
  geom_point(color = "#d95f02", size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Northeast Kingdom Enrollment Decline",
    subtitle = "Rural Vermont's population challenge intensifies",
    x = "Year",
    y = "Students"
  )
```

## High School Holds Steadier

While elementary enrollment has dropped sharply, high school grades have been more stable. Grade 12 enrollment has declined less than kindergarten, reflecting smaller incoming cohorts replacing larger graduating classes.

```{r hs-comparison-data}
hs_compare <- fetch_enr_multi(c(2010, 2015, 2020, 2024), use_cache = TRUE)

grade_compare <- hs_compare |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "12")) |>
  select(end_year, grade_level, n_students) |>
  mutate(grade_level = factor(grade_level, levels = c("K", "12"),
                              labels = c("Kindergarten", "Grade 12")))

grade_compare
```

```{r hs-comparison-chart, fig.cap="High school enrollment has declined less than kindergarten"}
ggplot(grade_compare, aes(x = end_year, y = n_students, color = grade_level)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Kindergarten" = "#1b9e77", "Grade 12" = "#7570b3")) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Kindergarten vs. Grade 12 Enrollment in Vermont",
    subtitle = "Elementary grades feel the demographic decline first",
    x = "Year",
    y = "Students",
    color = "Grade"
  )
```

## The Tiniest Schools in America

Vermont is home to some of the smallest public schools in the nation. Dozens of schools enroll fewer than 100 students, and some serve only a handful of children.

```{r tiny-schools-data}
tiny_schools <- enr_2024 |>
  filter(is_campus, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  filter(n_students > 0, n_students < 100) |>
  mutate(size_bin = cut(n_students, breaks = c(0, 25, 50, 75, 100),
                        labels = c("1-25", "26-50", "51-75", "76-99"))) |>
  count(size_bin)

tiny_schools
```

```{r tiny-schools-chart, fig.cap="Many Vermont schools serve fewer than 100 students"}
ggplot(tiny_schools, aes(x = size_bin, y = n)) +
  geom_col(fill = "#7570b3") +
  geom_text(aes(label = n), vjust = -0.5, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Vermont's Tiny Schools (Under 100 Students)",
    subtitle = "Small rural communities maintain neighborhood schools",
    x = "Enrollment Size",
    y = "Number of Schools"
  )
```

## The Chittenden Concentration

Chittenden County (the Burlington metro area) now enrolls a disproportionate share of Vermont's students. As rural areas shrink, the state's population concentrates in its one urban region.

```{r chittenden-data}
chittenden_trend <- fetch_enr_multi(c(2010, 2015, 2020, 2024), use_cache = TRUE)

# Identify Chittenden-area districts
chittenden_area <- chittenden_trend |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(region = if_else(
    grepl("Burlington|Essex|South Burlington|Winooski|Colchester|Chittenden",
          district_name, ignore.case = TRUE),
    "Chittenden County", "Rest of Vermont"
  )) |>
  group_by(end_year, region) |>
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

chittenden_area
```

```{r chittenden-chart, fig.cap="Chittenden County holds a growing share of Vermont students"}
ggplot(chittenden_area, aes(x = end_year, y = n_students, fill = region)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = c("Chittenden County" = "#006837",
                               "Rest of Vermont" = "#a1d99b")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Vermont Enrollment: Chittenden vs. Rest of State",
    subtitle = "Urban concentration increases as rural enrollment declines",
    x = "Year",
    y = "Students",
    fill = "Region"
  )
```

## Summary

Vermont's public school enrollment tells a story of rural challenges and demographic transition:

- **Steep decline**: Lost roughly a quarter of students since the early 2000s
- **Small scale**: Even the largest SUs serve fewer than 4,000 students
- **Supervisory structure**: Schools organized into SUs/SDs for shared administration
- **Chittenden dominance**: Burlington area anchors the state's enrollment
- **Many tiny districts**: Most SUs serve fewer than 1,000 students
- **Rural pressure**: Remote counties face the steepest enrollment declines

These patterns reflect Vermont's aging population, low birth rates, and the ongoing challenge of providing quality education in a small, rural state.
